A generalized errors-in-variables model, with application to single-cell CRISPR screens

### 2021-09-17 presentation/glmeiv_talk.tex
finish presentation draft

\begin{frame}
	\frametitle{Single-cell CRISPR screens entail sequencing gRNAs and mRNAs in individual cells.}

	\begin{figure}
		\centering
		\includegraphics[width=1\linewidth]{extra_figs/experimental_protocol}
	\end{figure}
\end{frame}

@@

As a simple example, suppose
$$
\begin{cases}
p_1, \dots, p_n \sim \textrm{Bern}(\pi) \\
y_i = \beta_m p_i + \ep_i \\
x_i = \beta_g p_i + \tau_i,
\end{cases}
$$
where $\ep_i \indep \tau_i$ and $\ep_i, \tau_i \sim N(0,1)$. We observe
$$ \{(x_1, y_1), \dots, (x_n, y_n)\},$$ and we want to estimate $\beta_m$ using the thresholding method. Assume $\beta_g$ and $\pi$ are known. We select the threshold $c$ so as to minimize the misclassification rate, i.e., $$c = \argmin_{c \in \R} \frac{1}{n} \sum_{i=1}^n \mathbb{I}(\hat{P}_i \neq P_i).$$

@@

\begin{frame}
\frametitle{Research question}
Does modeling the gRNA count distribution directly (thereby bypassing thresholding) improve estimation and inference?
\end{frame}

@@

\begin{frame}
\frametitle{Method summary}
\begin{itemize}
\item We extend the parametric model of \cite{Sarkar2021} to model both mRNA counts and gRNA counts in single-cell CRISPR screen experiments.
\begin{itemize}
\item Arbitrary exponential family distributions and link functions are supported. 
\end{itemize}
\item We (i) develop a fast EM algorithm to fit the model, (ii) derive an analytic expression for the observed information matrix, and (iii) implement a computational pipeline to deploy the method on HPC and cloud. 
\begin{itemize}
\item These features enable GLM-EIV to scale to large datasets.
\end{itemize}
\end{itemize}
\end{frame}

@@

@@

\begin{frame}
\frametitle{Real data analysis details}
\begin{itemize}
\item Quality control
\begin{itemize}
\item Lowly-expressed genes removed
\item Cells with library sizes below 5th percentile or above 95th percentile removed
\end{itemize}
\item mRNA model
\begin{itemize}
\item negative binomial distribution (with log link)
\item size parameter $\theta$ estimated from data.
\end{itemize}
\item gRNA model
\begin{itemize}
\item Poisson distribution (with log link) \cite{Choudhary2021}
\end{itemize}
\end{itemize}
\end{frame}

% However, the model gives us many insights into single-cell CRISPR screen datasets.
% What is the fold change in gRNA expression?
% What is the perturbation probability?

% Practical strategies:
% Is thresholding a reasonable strategy for these data?
% How can I choose a good threshold (cell-specific threshold; use only cells with high or low posterior membership probabilities, and throw all others out)?

>>>>

\begin{frame}
	\frametitle{Single-cell CRISPR screens entail sequencing gRNAs and mRNAs in individual cells.}
	\begin{figure}
		\centering
		\includegraphics[width=1\linewidth]{extra_figs/experimental_protocol}
	\end{figure}
\end{frame}

\begin{frame}
\frametitle{We investigate single-cell CRISPR screen datasets produced by \cite{Gasperini2019} and \cite{Xie2019}.}
\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{extra_figs/papers}
\end{figure}
\end{frame}

@@

As a simple example, suppose
$$
\begin{cases}
p_1, \dots, p_n \sim \textrm{Bern}(\pi) \\
y_i = \beta_m p_i + \ep_i \\
x_i = \beta_g p_i + \tau_i,
\end{cases}
$$
where $\ep_i \indep \tau_i$ and $\ep_i, \tau_i \sim N(0,1)$. We observe
$$ \{(x_1, y_1), \dots, (x_n, y_n)\},$$ and we want to estimate $\beta_m$ using the thresholding method. Assume $\beta_g$ and $\pi$ are known. We select the threshold $c$ so as to minimize the misclassification rate, i.e., $$c = \argmin_{c \in \R} \frac{1}{n} \sum_{i=1}^n \mathbb{I}(\hat{p}_i \neq p_i).$$

@@

\begin{frame}
\frametitle{Research question}
Does modeling the gRNA count distribution (thereby bypassing thresholding) improve estimation and inference?
\end{frame}

@@

\begin{frame}
\frametitle{Method summary}
\begin{itemize}
\item We extend the parametric model of \cite{Sarkar2021} to model both mRNA counts and gRNA counts.
\begin{itemize}
\item Arbitrary exponential family distributions and link functions are supported. 
\end{itemize}
\item We (i) propose a fast EM algorithm to fit the model, (ii) derive an analytic expression for the observed information matrix, and (iii) implement a computational pipeline to deploy the method on HPC and cloud. 
\begin{itemize}
\item These features enable GLM-EIV to scale to large datasets.
\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Simulation setup}
\begin{itemize}
\item No covariates
\item No offset terms (i.e., library size fixed at one)
\item Intercept terms fixed
\item $\pi$ fixed
\item $\beta_m$ fixed (and of moderate size)
\item $\beta_g$ varied over an interval
\item Gaussian, Poisson, and negative binomial distributions
\end{itemize}
\end{frame}

\begin{frame}
\begin{figure}
	\centering
	\includegraphics[width=1\linewidth]{../figures/fig3/simulation_warmup_arm_g}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{GLM-EIV outperforms the thresholding method on the simulated data for two main reasons:}
\begin{itemize}
\item[1.] GLM-EIV leverages information from \textit{both} modalities to assign perturbation identities to cells.
\item[2.] GLM-EIV generates \textit{soft} rather than \textit{hard} assignments, capturing the inherent uncertainty in whether a perturbation occurred.
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{We followed recommendations of \cite{Choudhary2021} for quality control and modeling.}

\begin{itemize}
\item Quality control
\begin{itemize}
\item Lowly-expressed genes filtered
\item Cells with library sizes below 5th percentile or above 95th percentile filtered
\end{itemize}
\item mRNA model
\begin{itemize}
\item used negative binomial distribution (with log link)
\item size parameter $\theta$ estimated from data
\end{itemize}
\item gRNA model
\begin{itemize}
\item used Poisson distribution (with log link)
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{The estimate $\hat{\beta}_g$ for $\beta_g$ was large on both datasets across all site types.}
\begin{center}
\textbf{Gasperini}
\end{center}
\begin{center}
\begin{tabular}{|c|c|}
	\hline 
	Site type & Mean $\exp\left(\beta_g\right)$ 95\% CI \\ 
	\hline 
 	 \textcolor{blue}{Candidate \textit{cis}} & $(4453, 5353)$ \\ 
	\hline 
	\textcolor{blue}{Negative control} &  $(4484, 5408)$ \\ 
	\hline
	\textcolor{blue}{TSS-targeting} & $(3605, 4205)$ \\
	\hline
\end{tabular} 
\end{center}

\begin{center}
	\textbf{Xie}
\end{center}
\begin{center}
	\begin{tabular}{|c|c|}
		\hline 
		Site type & Mean $\exp\left(\beta_g\right)$ 95\% CI \\ 
		\hline 
		\textcolor{blue}{Candidate \textit{cis}} & $(307, 324)$ \\ 
		\hline 
		\textcolor{blue}{Negative control} &  $(299, 316)$ \\
		\hline
	\end{tabular} 
\end{center}

\begin{frame}
\frametitle{GLM-EIV and thresholding exhibited similar CI coverage rates on the negative control pairs.}
\begin{center}
\begin{tabular}{|c|c|c|}
	\hline 
	Dataset & GLM-EIV & Thresholding \\ 
	\hline 
	\textcolor{blue}{Xie} & 93.7\% & 93.2\% \\ 
	\hline 
	\textcolor{blue}{Gasperini} & 90.6\% & 91.4\%  \\ 
	\hline 
\end{tabular}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Estimates and CIs produced by GLM-EIV and the thresholding method were similar.}
\begin{figure}
	\centering
	\includegraphics[width=1\linewidth]{../figures/fig5/comparison}
\end{figure}
\begin{center}
Candidate \textit{cis} pairs in Xie et al.\ data.
\end{center}
\end{frame}

\begin{frame}
\frametitle{We now can answer our core research question.}
\textbf{Research question}: Does modeling the gRNA count distribution (thereby bypassing thresholding) improve estimation and inference?
\begin{itemize}
\item Yes, if the problem is in a ``sufficiently challenging'' setting.
\item The real data that we analyzed, surprisingly, were in an ``easy'' setting.
\item Therefore, GLM-EIV and the thresholding method performed similarly on the real data.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{The proposed method could help solve practical challenges.}
\begin{itemize}
\item Selecting cell-specific thresholds
\item Identifying ``problem difficulty'' and thus whether thresholding is appropriate
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Together, GLM-EIV and SCEPTRE shed light on core analysis challenges posed by single-cell CRISPR screens, paving the way for the development of new methods.}
\begin{itemize}
	\item[1.] Perturbation unobserved
	\item[2.] Confounders and nuisance variables
	\item[3.] Possible model mispecification
	\item[4.] Background reads
	\item[5.] Highly discrete data
	\item[6.] Ineffective gRNAs
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Thank you.}
Acknowledgments:
\begin{itemize}
\item Xuran Wang helped with the Xie data preprocessing.
\item All analyses were run on Pittsburgh Supercomputer.
\end{itemize}
\end{frame}

% However, the model gives us many insights into single-cell CRISPR screen datasets.
% What is the fold change in gRNA expression?
% What is the perturbation probability?
% Practical strategies:
% Is thresholding a reasonable strategy for these data?
% How can I choose a good threshold (cell-specific threshold; use only cells with high or low posterior membership probabilities, and throw all others out)?
